Extended Data

### 2022-02-23 doc/manuscript/Dow_manuscript_Nature.Rmd
adjust language related to start/end of growing season
#113

<!--These following two paragraphs have a lot of overlap and need work. They should probably be merged into one. -->
In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014].
Current terrestrial ecosystem models, which have a rudimentary treatment of tree growth phenology[@delpierre_temperate_2016], effectively assume that longer growing seasons lead to increasing annual net CO~2~ uptake (i.e., net ecosystem exchange, _NEE_)[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014].
However, recent experimental and observational findings show that annual stem growth is not correlated with growing season length[@etzold_number_2022], but rather can be limited by sink factors[@keenan_timing_2015; @zani_increased_2020; @zohner_how_2021; @delpierre_wood_2016],<!--I (KAT) want to work on this--> and that positive effects of warm springs can be compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018]. 
These studies suggest that warmer spring temperatures may not have the expected positive effect on forest CO~2~ sequestration.

@@

Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the timing of stem growth but have little effect on total annual growth (Figs. `r schematic_summary`- `r tree_rings`). 
The observed advance in stem growth under warmer springs parallels phenological advances observed for canopy foliage (Fig. `r DOY_timing`a-b)[@keenan_net_2014;@finzi_carbon_2020;@stuble_plant_2021], which makes sense given close coordination of budburst and xylogenesis[@delpierre_temperate_2016], and _NEE_[@keenan_net_2014; @finzi_carbon_2020]. 
**However, inconsistent with the concept that an earlier start to growth would increase annual woody productivity, we demonstrate that warmer springs hasten the cessation of stem expansion and thereby have negligible effect on total annual growth for most species and locations (Fig. 3).**
**We demonstrate that warmer springs either have little effect on, or even hasten, the cessation of peak stem expansion and thereby have negligible effect on total annual stem growth for most species and locations (Fig. `r tree_rings`).** <!--repeated sentence-->
Our observations suggest that the cessation of rapid stem expansion, which occurs mid-summer near the time of <!--the summer solstice and--> peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020; @etzold_number_2022], is likely driven by cues other than photosynthate limitation, such as water stress, daylength, or sink limitation[@zweifel_why_2021; @etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;@fu_variation_2014; @delpierre_wood_2016; @delpierre_temperate_2016]. <!--note that the timing of the cessation of PRIMARY shoot growth is linked to bud-set, which must be accomplished before the first frost, and has been the subject of some research (see delpierre_temperate_2016, rohde_temperature_2011), but I don't know what if any linkage exists between secondary growth and bud set.-->
Our tree-ring analysis (Fig. `r tree_rings`) demonstrates that the primary effect of warming temperatures on annual ring-width is not an augmentation through an earlier start to growth, but rather a reduction associated to drought stress during the peak growing season[@dorangeville_drought_2018;@etzold_number_2022]. 
Warm springs may also amplify summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018;@zhang_drought_2021]. 
However, we did not find evidence that higher spring temperatures were correlated with summer drought stress in our dendrometer band analysis, implying that the effects of warm spring temperatures on stem growth elucidated here (Fig. `r schematic_summary`) were not attributable to summer drought.

Our finding that interannual variation in stem growth is more strongly linked to conditions during the peak growing season than to growing season length aligns with recent studies on annual stem growth[@etzold_number_2022; @tumajer_limitation_2022;@delpierre_wood_2016] and with parallel findings for _NEE_[@zhou_explaining_2016; @fu_maximum_2019].
However, there is also a disconnect with findings that _NEE_ is at least modestly greater in years with warm springs[@keenan_net_2014] or long growing seasons[@zhou_explaining_2016; @fu_maximum_2019;@finzi_carbon_2020].
Warming advances spring phenology and may advance or delay autumn senescence depending on timing of warming and water availability[@zohner_ongoing_2019;@xie_predicting_2018; @crabbe_extreme_2016], with delays more common across eastern North America [@buermann_widespread_2018;@finzi_carbon_2020;@keenan_net_2014], implying that warming temperatures are lengthening the period from peak stem growth to the cessation of CO~2~ uptake by the ecosystem.
We show that the extra C fixation in years with warm springs does not substantially augment stem growth, but it remains unclear how it is allocated within the ecosystem.
There are two main possibilities, which hold contrasting implications for the response of forest C balance to rising spring temperatures.

@@

A second possibility is that any additional C fixed during years with warm springs may be allocated to plant functions other than stem growth, including respiration, NSC production, reproduction, and production of foliage, roots[@zohner_how_2021], or root exudates. 
Much of this C would have a relatively short residence time within the ecosystem, and C loss through fall or winter respiration may offset gains from an earlier spring[@oishi_warmer_2018; @buermann_widespread_2018].
However, C allocated to nonstructural carbohydrates or relatively short-lived plant tissues would typically remain in the ecosystem beyond the end of the year[@trumbore_quantifying_2002], such that the long-term effect of warm springs on the forest C balance would not be captured in analyses of interannual variation[@zhou_explaining_2016; @fu_maximum_2019;@keenan_net_2014].
Studies within or including the temperate deciduous biome that examined long-term trends in growing season length and ecosystem C uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@finzi_carbon_2020] -- as opposed to their interannual variation -- showed increasing trends in both variables, suggesting that the C not allocated to woody productivity within the current year has a multi-year residence time within the ecosystem.
However, given our finding that warm springs do not significantly enhance woody productivity, this C is likely to have a relatively short residence time within the ecosystem.

>>>>

<!--These following two paragraphs have a lot of overlap and need work. They should probably be merged into one. -->
In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014].
Current terrestrial ecosystem models, which have a rudimentary treatment of tree growth phenology[@delpierre_temperate_2016], effectively assume that longer growing seasons lead to increasing annual net CO~2~ uptake (i.e., net ecosystem exchange, _NEE_)[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014].
However, recent experimental and observational findings show that annual stem growth is not correlated with growing season length[@etzold_number_2022], but rather can be limited by sink factors[@keenan_timing_2015; @zani_increased_2020; @zohner_how_2021; @delpierre_wood_2016],<!--I (KAT) want to work on this (and check if keenan_timing_2015 looks at stem growth)--> and that positive effects of warm springs can be compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018]. 
These studies suggest that warmer spring temperatures may not have the expected positive effect on forest CO~2~ sequestration.

@@

Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the timing of stem growth but have little effect on total annual growth (Figs. `r schematic_summary`- `r tree_rings`). 
The observed advance in stem growth under warmer springs parallels phenological advances observed for canopy foliage (Fig. `r DOY_timing`a-b)[@keenan_net_2014;@finzi_carbon_2020;@stuble_plant_2021], which makes sense given close coordination of budburst and xylogenesis[@delpierre_temperate_2016], and _NEE_[@keenan_net_2014; @finzi_carbon_2020]. 
**However, inconsistent with the concept that an earlier start to growth would increase annual woody productivity, we demonstrate that warmer springs hasten the deceleration of stem expansion and thereby have negligible effect on total annual growth for most species and locations (Fig. 3).**
**We demonstrate that warmer springs either have little effect on, or even hasten, the deceleration of stem expansion and thereby have negligible effect on total annual stem growth for most species and locations (Fig. `r tree_rings`).** <!--repeated sentence-->
Our observations suggest that the deceleration of stem expansion (starting on $DOY_{g_{max}}$), which occurs mid-summer near the time of <!--the summer solstice and--> peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020; @etzold_number_2022], is likely driven by cues other than photosynthate limitation, such as water stress, daylength, or sink limitation[@zweifel_why_2021; @etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;@fu_variation_2014; @delpierre_wood_2016; @delpierre_temperate_2016]. <!--note that the timing of the cessation of PRIMARY shoot growth is linked to bud-set, which must be accomplished before the first frost, and has been the subject of some research (see delpierre_temperate_2016, rohde_temperature_2011), but I don't know what if any linkage exists between secondary growth and bud set.-->
Our tree-ring analysis (Fig. `r tree_rings`) demonstrates that the primary effect of warming temperatures on annual ring-width is not an augmentation through an earlier start to growth, but rather a reduction associated to drought stress during the peak growing season[@dorangeville_drought_2018;@etzold_number_2022]. 
Warm springs may also amplify summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018;@zhang_drought_2021]. 
However, we did not find evidence that higher spring temperatures were correlated with summer drought stress in our dendrometer band analysis, implying that the effects of warm spring temperatures on stem growth elucidated here (Fig. `r schematic_summary`) were not attributable to summer drought.

Our finding that interannual variation in stem growth is more strongly linked to conditions during the peak growing season than to the length of the season of active stem growth aligns with recent studies on annual stem growth[@etzold_number_2022; @tumajer_limitation_2022;@delpierre_wood_2016] and with parallel findings for _NEE_[@zhou_explaining_2016; @fu_maximum_2019].
However, there is also a disconnect with findings that _NEE_ is at least modestly greater in years with warm springs[@keenan_net_2014] or long growing seasons[@zhou_explaining_2016; @fu_maximum_2019;@finzi_carbon_2020].
Warming advances spring phenology and may advance or delay autumn senescence depending on timing of warming and water availability[@zohner_ongoing_2019;@xie_predicting_2018; @crabbe_extreme_2016], with delays more common across eastern North America [@buermann_widespread_2018;@finzi_carbon_2020;@keenan_net_2014], implying that warming temperatures are lengthening the period from peak stem growth to the cessation of CO~2~ uptake by the ecosystem.
We show that the extra C fixation in years with warm springs does not substantially augment stem growth, but it remains unclear how it is allocated within the ecosystem.
There are two main possibilities, which hold contrasting implications for the response of forest C balance to rising spring temperatures.

@@

A second possibility is that any additional C fixed during years with warm springs may be allocated to plant functions other than stem growth, including respiration, NSC storage, reproduction, and production of foliage, roots[@zohner_how_2021], or root exudates. 
Much of this C would have a relatively short residence time within the ecosystem, and C loss through fall or winter respiration may offset gains from an earlier spring[@oishi_warmer_2018; @buermann_widespread_2018].
However, C allocated to nonstructural carbohydrates or relatively short-lived plant tissues would typically remain in the ecosystem beyond the end of the year[@trumbore_quantifying_2002], such that the long-term effect of warm springs on the forest C balance would not be captured in analyses of interannual variation[@zhou_explaining_2016; @fu_maximum_2019;@keenan_net_2014].
Studies within or including the temperate deciduous biome that examined long-term trends in growing season length and ecosystem C uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;@finzi_carbon_2020] -- as opposed to their interannual variation -- showed increasing trends in both variables, suggesting that the C not allocated to woody productivity within the current year has a multi-year residence time within the ecosystem.
However, given our finding that warm springs do not significantly enhance woody productivity, this C is likely to have a relatively short residence time within the ecosystem.
